---
title: "Analysis of the ToothGrowth data in R"
author: "Rinnette Ramdhanie"
date: "13 November 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loadDataLib, echo = FALSE, warning = FALSE, message = FALSE}
        data(ToothGrowth)        
        library(ggplot2)

```
## Overview




##Exploratory Analysis
The structure of the dataframe was checked as well as the different doses adminstered and the first few rows of data.

```{r summary}
        str(ToothGrowth)
        unique(ToothGrowth$dose)
        head(ToothGrowth)
```

The data consists of 60 rows and 3 variables: len, supp and dose.

* len: length of the odontoplasts in 60 guinea pigs
* supp: delivery method which can be administered using one of two supplements - orange juice (OJ) and ascorbic acid (VC)
* dose: three doses, 0.5, 1.0, and 2.0 milligrams per day, were administered for each delivery method

```{r plotData}
        ggplot(data = ToothGrowth, aes(x = supp, y = len)) +
                geom_boxplot(aes(fill = supp)) +
                facet_wrap(~ dose) +
                labs(title = "Tooth growth by dosage and type of supplement", 
                     x = "Type of Supplement", y = "Length of teeth")

```

Observations:

* Generally orange juice was a better source of Vitamin C causing the odontoplasts to grow longer for doses 0.5 and 1.
* A higher dose using either delivery method increased growth
* and some more stuff ... 

## Tests
### Comparing effects of supplements on tooth growth
Let $\mu_{oj}$ be the mean of tooth length of guinea pigs who was given orange juice and $\mu_{vc}$ the mean of tooth length of guinea pigs given ascorbic acid (vitamin C).

        $H_{0}: \mu_{oj}-\mu_{vc} = 0$
        $H_{a}: \mu_{oj}-\mu_{vc} > 0$
        
        Testing on a 5% significance level ie. alpha is 0.05.
```{r}
        oj <- ToothGrowth$len[ToothGrowth$supp == 'OJ']
        vc <- ToothGrowth$len[ToothGrowth$supp == 'VC']
        t.test(oj, vc, paired = FALSE, var.equal = FALSE, alternative = "greater")$p.value
```

Since the p-value is 0.03032 which is less than alpha, 0.05, $H_{0}$ is rejected in favor of $H_{a}$.  This means that there is a definite difference in tooth growth between pigs administered with orange juice and those administered with vitamin C.

### Comparing effects of dosage on tooth growth




## Results


## Conclusion

Assumptions go here

