---
title: "Analysis of the ToothGrowth data in R"
author: "Rinnette Ramdhanie"
date: "13 November 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loadDataLib, warning = FALSE, message = FALSE}
        data(ToothGrowth)        
        library(ggplot2)

```
## Overview




#Exploratory Analysis


```{r summary}
        str(ToothGrowth)
        unique(ToothGrowth$dose)
        head(ToothGrowth)
```

The data consists of 60 rows and 3 variables: len, supp and dose.

* len: length of the odontoplasts in 60 guinea pigs
* supp: delivery method which can be one of two - orange juice (OJ) and ascorbic acid (VC)
* dose: three doses - 0.5, 1.0, or 2.0

```{r plotData}
        ggplot(data = ToothGrowth, aes(x = supp, y = len)) +
                geom_boxplot(aes(fill = supp)) +
                facet_wrap(~ dose)

```

Observations:

* Generally orange juice was a better source of Vitamin C causing the odontoplasts to grow longer for doses 0.5 and 1.
* A higher dose using either delivery method increased growth
* and some more stuff ... 

## Tests
 

## Results


## Conclusion

Assumptions go here

