---
title: "Analysis of the ToothGrowth data in R"
author: "Rinnette Ramdhanie"
date: "13 November 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r loadDataLib, echo = FALSE, warning = FALSE, message = FALSE}
        # Load data and libraries        
                data(ToothGrowth)        
                library(ggplot2)
                library(dplyr)
```
## Overview




##Exploratory Analysis
The structure of the dataframe was checked as well as the different doses adminstered and the first few rows of data.

```{r summary, echo = TRUE}
        # Summary of data        
        summary(ToothGrowth)
        str(ToothGrowth)
        unique(ToothGrowth$dose)
        head(ToothGrowth, 3)
```

The data consists of 60 rows and 3 variables: len, supp and dose.

* len: length of the odontoplasts in 60 guinea pigs
* supp: delivery method which can be administered using one of two supplements - orange juice (OJ) and ascorbic acid (VC)
* dose: three doses, 0.5, 1.0, and 2.0 milligrams per day, were administered for each delivery method

```{r plotData, fig.width=7,fig.height=3}
        # Boxplot of teeth length against supplement for each dose
        ggplot(data = ToothGrowth, aes(x = supp, y = len)) +
                geom_boxplot(aes(fill = supp)) +
                facet_wrap(~ dose) +
                labs(title = "Tooth growth by dosage and type of supplement", 
                     x = "Type of Supplement", y = "Length of teeth")

```

Initial Observations:

* Generally it seems like orange juice was a better source of Vitamin C causing the odontoplasts to grow longer for doses of 0.5 mg and 1.0 mg.
* A higher dose using either delivery method increased growth.

## Tests
All testing was done using a 5% significance level ie. alpha is 0.05.  The p-values are displayed for each test.

### Comparing effects of supplements on tooth growth
Let $\mu_{oj}$ be the mean of tooth length of guinea pigs who was given orange juice and $\mu_{vc}$ the mean of tooth length of guinea pigs given ascorbic acid (vitamin C) ie. 
        $H_{0}: \mu_{oj} - \mu_{vc} = 0$
        $H_{a}: \mu_{oj} - \mu_{vc} > 0$

```{r}
        oj <- ToothGrowth$len[ToothGrowth$supp == 'OJ']
        vc <- ToothGrowth$len[ToothGrowth$supp == 'VC']
        t.test(oj, vc, paired = FALSE, var.equal = FALSE, alternative = "greater")$p.value
```

Since the p-value is 0.03032 which is less than alpha, 0.05, $H_{0}$ is rejected in favor of $H_{a}$.  This means that there is storng evidence to support a difference in tooth growth between guinea pigs administered with orange juice and those administered with vitamin C.

### Comparing effects of dosage on tooth growth

For each of the supplements the following was tested:

* difference in tooth growth between dose 0.5 mg and 1.0 mg
* difference in tooth growth between dose 1.0 mg and 2.0 mg


#### Comparing 0.5mg dosage with 1.0 mg of orange juice

Let $\mu_{0.5}$ be the mean of tooth length of guinea pigs who was give 0.5 mg of the supplement and $\mu_{1.0}$ the mean of tooth length of guinea pigs given 1.0 mg of the supplement. 
        $H_{0}: \mu_{1.0}-\mu_{0.5} = 0$
        $H_{a}: \mu_{1.0}-\mu_{0.5} > 0$
        
```{r echo = TRUE}
        g1 <- filter(ToothGrowth, supp == "OJ", dose == 0.5)$len
        g2 <- filter(ToothGrowth, supp == "OJ", dose == 1.0)$len
        t.test(g2, g1, paired = FALSE, var.equal = FALSE, alternative = "greater")$p.value
```

Since the p-value is 4.39246e-05 which is less than alpha, 0.05, $H_{0}$ is rejected in favor of $H_{a}$.  This means that there is a definite difference in tooth growth between guinea pigs administered with 0.5 mg orange juice and those administered with 1.0 mg vitamin C.


#### Comparing 1.0 mg dosage with 2.0 mg of orange juice
Let $\mu_{1.0}$ be the mean of tooth length of guinea pigs who was give 1.0 mg of the supplement and $\mu_{2.0}$ the mean of tooth length of guinea pigs given 1.0 mg of the supplement. 
        $H_{0}: \mu_{2.0} - \mu_{1.0} = 0$
        $H_{a}: \mu_{2.0} - \mu_{1.0} > 0$
        
```{r echo = TRUE}
        g1 <- filter(ToothGrowth, supp == "OJ", dose == 1.0)$len
        g2 <- filter(ToothGrowth, supp == "OJ", dose == 2.0)$len
        t.test(g2, g1, paired = FALSE, var.equal = FALSE, alternative = "greater")$p.value
```

Since the p-value is 0.0196 which is less than alpha, 0.05, $H_{0}$ is rejected in favor of $H_{a}$.  This means that there is a definite difference in tooth growth between guinea pigs administered with 1.0 mg of orange juice and those administered with 2.0 mg.

#### Comparing 0.5mg dosage with 1.0 mg of vitamin C

Let $\mu_{0.5}$ be the mean of tooth length of guinea pigs who was give 0.5 mg of the supplement and $\mu_{1.0}$ the mean of tooth length of guinea pigs given 1.0 mg of the supplement. 
        $H_{0}: \mu_{1.0} - \mu_{0.5} = 0$
        $H_{a}: \mu_{1.0} - \mu_{0.5} > 0$
        
```{r echo = TRUE}
        g1 <- filter(ToothGrowth, supp == "VC", dose == 0.5)$len
        g2 <- filter(ToothGrowth, supp == "VC", dose == 1.0)$len
        t.test(g2, g1, paired = FALSE, var.equal = FALSE, alternative = "greater")$p.value
```

Since the p-value is 3.41e-07 which is much less than alpha, 0.05, $H_{0}$ is rejected in favor of $H_{a}$.  This means that there is very strong evidence to support a difference in tooth growth between guinea pigs administered with 0.5 mg of vitamin C and those administered with 1.0 mg vitamin C.


#### Comparing 1.0 mg dosage with 2.0 mg of orange juice
Let $\mu_{1.0}$ be the mean of tooth length of guinea pigs who was give 1.0 mg of the supplement and $\mu_{2.0}$ the mean of tooth length of guinea pigs given 1.0 mg of the supplement. 
        $H_{0}: \mu_{2.0} - \mu_{1.0} = 0$
        $H_{a}: \mu_{2.0} - \mu_{1.0} > 0$
        
```{r echo = TRUE}
        g1 <- filter(ToothGrowth, supp == "VC", dose == 1.0)$len
        g2 <- filter(ToothGrowth, supp == "VC", dose == 2.0)$len
        t.test(g2, g1, paired = FALSE, var.equal = FALSE, alternative = "greater")$p.value
```

Since the p-value is 4.58e-05 which is much less than alpha, 0.05, $H_{0}$ is rejected in favor of $H_{a}$.  This means that there is very strong evidence to support a difference in tooth growth between guinea pigs administered with 1.0 mg of vitamin C and those administered with 2.0 mg vitamin C.

## Results


## Conclusion

Assumptions go here

## Appendix
```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```