---
title: "Analysis of the ToothGrowth data in R"
author: "Rinnette Ramdhanie"
date: "13 November 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r loadDataLib,  warning = FALSE, message = FALSE}
        # Load data and libraries        
                data(ToothGrowth)        
                library(ggplot2)
                library(dplyr)
```
## Overview
The ToothGrowth data in the R datasets package was analysed in this report. Some exploratory analysis was done on the data, then hypothesis testing was used to compare tooth growth by dosage and supplement administered. The testing showed that orange juice has a greater effect on tooth growth than vitamin C. Also larger doses of any supplement had a greater effect on tooth growth.



## Exploratory Analysis
A summary of the data and the structure of the dataframe were checked as well as the different doses administered and the first few rows of data.

```{r summary}
        # Summary of data        
        summary(ToothGrowth)
        str(ToothGrowth)
        unique(ToothGrowth$dose)
        head(ToothGrowth, 3)
```

The data consists of 60 rows and 3 variables: len, supp and dose.

* len: length of the odontoplasts in 60 guinea pigs
* supp: delivery method which can be administered using one of two supplements - orange juice or ascorbic acid (vitamin C).
* dose: three doses, 0.5, 1.0, and 2.0 milligrams per day, were administered for each delivery method

The figure below shows a boxplot of teeth length against type of supplement for each dosage.

Initial observations:

* Orange juice seems to have increased teeth growth at doses 0.5 and 1.0 mg. At 2.0 mg the vitamin C seemed to have affected teeth growth more than the orange juice.
* In  general a higher dose using either delivery method increased growth.

```{r plotData, fig.width=7,fig.height=3}
        # Boxplot of teeth length against supplement for each dose
        ggplot(data = ToothGrowth, aes(x = supp, y = len)) +
                geom_boxplot(aes(fill = supp)) +
                facet_wrap(~ dose) +
                labs(title = "Tooth growth by dosage and type of supplement", 
                     x = "Type of Supplement", y = "Length of teeth")
```

## Tests
All tests were done using a 5% significance level ie. alpha is 0.05.  The results are displayed for each test.

### Comparing the effect of supplement type on tooth growth
*Test 1*

Let $\mu_{oj}$ be the mean tooth length for guinea pigs that were given orange juice and $\mu_{vc}$ the mean tooth length of guinea pigs given ascorbic acid. The following hypotheses were tested: $H_{0}: \mu_{oj} - \mu_{vc} = 0$ and $H_{a}: \mu_{oj} - \mu_{vc} > 0$.

\scriptsize
```{r}
        # Comparing effects of supplement type on tooth growth
                oj <- filter(ToothGrowth, supp == 'OJ')$len
                vc <- filter(ToothGrowth, supp == 'VC')$len
                t.test(oj, vc, alternative = "greater")
```
\normalsize

The results of the test show that $H_0$ was rejected in favor of $H_a$ and the "true difference in means is greater than zero".  Therefore orange juice has a statistically significant effect on tooth growth than vitamin C.

### Comparing the effect of dosage on tooth growth

For each of the supplements the following were tested:

* difference in tooth growth between doses 0.5 mg and 1.0 mg (Tests 2 and 4)
* difference in tooth growth between doses 1.0 mg and 2.0 mg (Tests 3 and 5)

The p-value for each test is shown.  If $p.value < 0.05$ then this proves that there is a statistical difference between the two means and $H_0$ is rejected in favour of $H_a$.

*Test 2: Comparing 0.5 mg with 1.0 mg of orange juice*

Let $\mu_{0.5}$ be the mean tooth length of guinea pigs that were given 0.5 mg of orange juice and $\mu_{1.0}$ the mean tooth length of guinea pigs given 1.0 mg of orange juice. The following hypotheses were tested: $H_{0}: \mu_{1.0}-\mu_{0.5} = 0$ and $H_{a}: \mu_{1.0}-\mu_{0.5} > 0$.

```{r}
        g1 <- filter(ToothGrowth, supp == "OJ", dose == 0.5)$len
        g2 <- filter(ToothGrowth, supp == "OJ", dose == 1.0)$len
        t.test(g2, g1, alternative = "greater")$p.value
```


*Test 3: Comparing 1.0 mg with 2.0 mg of orange juice*

Let $\mu_{1.0}$ be the mean tooth length of guinea pigs that were given 1.0 mg of orange juice and $\mu_{2.0}$ the mean tooth length of guinea pigs given 2.0 mg of orange juice. The following hypotheses were tested:   $H_{0}: \mu_{2.0} - \mu_{1.0} = 0$ and $H_{a}: \mu_{2.0} - \mu_{1.0} > 0$.
        
```{r}
        g1 <- filter(ToothGrowth, supp == "OJ", dose == 1.0)$len
        g2 <- filter(ToothGrowth, supp == "OJ", dose == 2.0)$len
        t.test(g2, g1, alternative = "greater")$p.value
```


*Test 4: Comparing 0.5mg with 1.0 mg of vitamin C*

Let $\mu_{0.5}$ be the mean tooth length of guinea pigs that were given 0.5 mg of vitamin C and $\mu_{1.0}$ the mean tooth length of guinea pigs given 1.0 mg of vitamin C. The following hypotheses were tested: $H_{0}: \mu_{1.0} - \mu_{0.5} = 0$ and $H_{a}: \mu_{1.0} - \mu_{0.5} > 0$.
        
```{r}
        g1 <- filter(ToothGrowth, supp == "VC", dose == 0.5)$len
        g2 <- filter(ToothGrowth, supp == "VC", dose == 1.0)$len
        t.test(g2, g1, alternative = "greater")$p.value
```


*Test 5: Comparing 1.0 mg with 2.0 mg of orange juice*
Let $\mu_{1.0}$ be the mean tooth length of guinea pigs that were given 1.0 mg of vitamin C and $\mu_{2.0}$ the mean tooth length of guinea pigs given 1.0 mg of vitamin C. The following hypotheses were tested: $H_{0}: \mu_{2.0} - \mu_{1.0} = 0$ and $H_{a}: \mu_{2.0} - \mu_{1.0} > 0$.
        
```{r}
        g1 <- filter(ToothGrowth, supp == "VC", dose == 1.0)$len
        g2 <- filter(ToothGrowth, supp == "VC", dose == 2.0)$len
        t.test(g2, g1, alternative = "greater")$p.value
```


## Results
In the first test $H_0$ was rejected in favour of $H_a$.  In the remaining tests, the p-values were less tha 0.05 therefore supporting the alternative hypotheses, that the true difference in means is not equal to 0.  


## Conclusion
It was assumed that the data is normally distributed and that the groups used in the testing were independent, normally distributed and had unequal variances.

The first test showed that since $\mu_{oj} > \mu_{vc}$, orange juice has a statistically more significant effect than vitamin C on tooth growth.  Similarly, in the remaining tests, since $\mu_{larger dose} > \mu_{smaller dose}$ in each case no matter the supplement used, it means that statistically, a larger dose of either of the two supplements will will have a significantly greater effect on tooth growth compared to a smaller dose.

## Appendix
```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```